---
title: "Understanding Linear Regression"
subtitle: "From Baseline Averages to Best-Fit Lines with Galton's Data"
format:
  revealjs:
    transition: fade
    slide-number: true
    code-fold: true
execute:
  echo: false
  warning: false
---

```{r}
#| label: setup
library(ggplot2)
library(dplyr)
library(HistData)
library(flextable)
library(tibble)

col_point   <- "#0072B2"
col_mean    <- "#E69F00"
col_naive   <- "#D55E00"
col_ols     <- "#009E73"
col_predict <- "#CC79A7"

theme_deck <- theme_minimal(base_size = 16) +
  theme(plot.title = element_text(face = "bold"),
        plot.subtitle = element_text(color = "#7A7A7A"))
```

```{r}
#| label: load-data
data("GaltonFamilies", package = "HistData")

df <- GaltonFamilies %>%
  filter(gender == "male") %>%
  select(father, childHeight) %>%
  rename(son = childHeight) %>%
  mutate(father = father * 2.54, son = son * 2.54)

n_sons      <- nrow(df)
mean_son    <- mean(df$son)
mean_father <- mean(df$father)
future_fathers <- c(172.5, 185.0, 163.2)
known_data <- df %>% head(6) %>%
  mutate(father = round(father,1), son = round(son,1))
```

## Dataset

Galton's 1886 study: heights of `r n_sons` father-son pairs.

Everything on these slides is computed live from this data.

## Hypothesis

-   Can a father's height predict his son's height?

## The Outcome Variable

Just the son's heights, no predictor yet:

```{r}
known_data %>% select(son) %>% flextable() %>%
  set_header_labels(son = "Son's Height (cm)") %>%
  align(align = "center", part = "all") %>% autofit()
```

## Question

If this is all we know, what's your best guess for a **new** boy's height?

## Answer

The **average** — it won't be perfect for any one boy, but it's the best single number we have.

## Average Height

$$\text{mean} = `r round(mean_son,1)`\text{ cm}$$

```{r}
ggplot(df, aes(x = son)) +
  geom_histogram(bins = 15, fill = col_point, color = "white", alpha = 0.85) +
  geom_vline(xintercept = mean_son, color = col_mean, linewidth = 1.2, linetype = "dashed") +
  labs(title = "Distribution of Son Heights", x = "Son's Height (cm)", y = "Count") +
  theme_deck
```

## 2 Variables

Now we're also given the father's height:

```{r}
future_data <- tibble(father = as.character(future_fathers), son = rep("?", 3))
bind_rows(known_data %>% mutate(father = as.character(father), son = as.character(son)), future_data) %>%
  flextable() %>%
  set_header_labels(father = "Father (cm)", son = "Son (cm)") %>%
  align(align = "center", part = "all") %>%
  color(i = 7:9, j = 2, color = col_predict) %>%
  bold(i = 7:9, j = 1:2) %>% autofit()
```

Can we do better than just guessing the mean?

## Plotting the 2 Variables

```{r}
ggplot(df, aes(x = father, y = son)) +
  geom_point(color = col_point, alpha = 0.5, size = 2) +
  labs(title = "Father's Height vs. Son's Height", x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

## Approach 1: Ignore the Father

Keep guessing the mean, no matter what. Every prediction is the same.

## Approach 2: Father = Son

Predict the son will be exactly as tall as his father.

## A Middle Ground?

-   Mean-only: father matters **0%**.
-   Y = X: father matters **100%**.

Reality is somewhere in between — let's find it.

## The Equation

$$y = \text{intercept} + (\text{slope} \times x)$$

Every straight-line model is just a choice of these two numbers.

## Two Questions Every Line Must Answer

1.  Where does it cross the y-axis? (**intercept**)
2.  How steep is it? (**slope**)

## Mean-Only Model: The Slope

The model ignores $x$ completely, so:

$$\text{slope} = 0$$

## Mean-Only Model: The Intercept

```{r}
slope_mean <- 0
intercept_mean <- mean(df$son) - slope_mean * mean(df$father)
```

$$\text{intercept} = \text{mean}(y) = `r round(intercept_mean,1)`$$

## Mean-Only Model: The Line

$$y = `r round(intercept_mean,1)` + (0 \times x) = `r round(intercept_mean,1)`$$

A flat line — same prediction for every father.

## Mean-Only Model: Residuals

::: {.panel-tabset}

### Zoomed

```{r}
df_pred <- df %>% mutate(pred_mean = intercept_mean)
ggplot(df_pred, aes(x = father, y = son)) +
  geom_segment(aes(xend = father, yend = pred_mean), color = col_mean, linewidth = 0.15, alpha = 0.3) +
  geom_point(color = col_point, alpha = 0.4, size = 1.6) +
  geom_hline(yintercept = intercept_mean, color = col_mean, linewidth = 1.1, linetype = "dashed") +
  labs(title = "Mean-Only Model", x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

### True Origin

```{r}
ggplot(df_pred, aes(x = father, y = son)) +
  geom_hline(yintercept = intercept_mean, color = col_mean, linewidth = 1.1, linetype = "dashed") +
  geom_point(color = col_point, alpha = 0.4, size = 1.6) +
  geom_point(aes(x = 0, y = intercept_mean), color = "red", size = 3) +
  annotate("text", x = 5, y = intercept_mean + 8, label = "intercept: (0, mean)", color = "red", hjust = 0) +
  expand_limits(x = 0, y = 0) +
  labs(title = "Mean-Only Model - Full Axes to Origin", x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

:::

## Naive Model: The Slope

We want $y$ to increase by 1 for every 1-unit increase in $x$:

$$\text{slope} = 1$$

A chosen rule, not derived from the data.

## Naive Model: The Intercept

We want the line to pass exactly through $y = x$:

$$\text{intercept} = 0$$

*(Note: this isn't the "best-fit" intercept for slope = 1 — it's simply the rule "son = father.")*

## Naive Model: The Line

$$y = 0 + (1 \times x) = x$$

## Naive Model: Residuals

::: {.panel-tabset}

### Zoomed

```{r}
df_pred <- df_pred %>% mutate(pred_naive = father)
ggplot(df_pred, aes(x = father, y = son)) +
  geom_segment(aes(xend = father, yend = pred_naive), color = col_naive, linewidth = 0.15, alpha = 0.3) +
  geom_point(color = col_point, alpha = 0.4, size = 1.6) +
  geom_abline(intercept = 0, slope = 1, color = col_naive, linewidth = 1.1, linetype = "dashed") +
  labs(title = "Naive Model (Y = X)", x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

### True Origin

```{r}
ggplot(df_pred, aes(x = father, y = son)) +
  geom_abline(intercept = 0, slope = 1, color = col_naive, linewidth = 1.1, linetype = "dashed") +
  geom_point(color = col_point, alpha = 0.4, size = 1.6) +
  geom_point(aes(x = 0, y = 0), color = "red", size = 3) +
  annotate("text", x = 5, y = 15, label = "intercept: (0, 0)", color = "red", hjust = 0) +
  expand_limits(x = 0, y = 0) +
  labs(title = "Naive Model - Full Axes to Origin", x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

:::

## Why Not Just Pick One?

-   Mean-only throws away the father's height entirely.
-   Y = X throws away everything we know about the whole population.

We want the line that makes the **smallest possible errors** — let's solve for it.

## OLS: The Slope

$$\text{slope} = \frac{\text{covariance}(x,y)}{\text{variance}(x)}$$

```{r}
slope_ols <- cov(df$father, df$son) / var(df$father)
```

$$\text{slope} = `r round(slope_ols,3)`$$

## OLS: The Intercept

$$\text{intercept} = \text{mean}(y) - (\text{slope} \times \text{mean}(x))$$

```{r}
intercept_ols <- mean(df$son) - slope_ols * mean(df$father)
ols_full <- lm(son ~ father, data = df)
```

$$\text{intercept} = `r round(intercept_ols,1)`$$

## OLS: The Line

$$y = `r round(intercept_ols,1)` + (`r round(slope_ols,3)` \times x)$$

This line is **solved for**, not guessed — it's the one that minimizes total prediction error.

## OLS Regression Line

::: {.panel-tabset}

### Zoomed

```{r}
df_full <- df %>% mutate(ols_pred = predict(ols_full))
ggplot(df_full, aes(x = father, y = son)) +
  geom_segment(aes(xend = father, yend = ols_pred), color = col_ols, linewidth = 0.15, alpha = 0.3) +
  geom_point(color = col_point, alpha = 0.4, size = 1.6) +
  geom_abline(intercept = intercept_ols, slope = slope_ols, color = col_ols, linewidth = 1.1) +
  labs(title = "OLS Linear Regression", subtitle = paste("N =", n_sons, "sons"),
       x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

### True Origin

```{r}
ggplot(df_full, aes(x = father, y = son)) +
  geom_abline(intercept = intercept_ols, slope = slope_ols, color = col_ols, linewidth = 1.1) +
  geom_point(color = col_point, alpha = 0.4, size = 1.6) +
  geom_point(aes(x = 0, y = intercept_ols), color = "red", size = 3) +
  annotate("text", x = 5, y = intercept_ols + 8, label = "intercept", color = "red", hjust = 0) +
  expand_limits(x = 0, y = 0) +
  labs(title = "OLS Linear Regression - Full Axes to Origin",
       subtitle = paste("N =", n_sons, "sons"), x = "Father (cm)", y = "Son (cm)") +
  theme_deck
```

:::

## Comparing the Three Models

Every model makes prediction errors. We can measure how far off each one typically is — we'll cover exactly how in a future class.

```{r}
df_eval <- df %>%
  mutate(pred_mean = intercept_mean, pred_naive = father, pred_ols = predict(ols_full))

err <- tibble(
  Model = c("Mean-only", "Naive (Y=X)", "OLS"),
  Error = c(
    sqrt(mean((df_eval$son - df_eval$pred_mean)^2)),
    sqrt(mean((df_eval$son - df_eval$pred_naive)^2)),
    sqrt(mean((df_eval$son - df_eval$pred_ols)^2))
  )
)

ggplot(err, aes(x = reorder(Model, Error), y = Error, fill = Model)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  scale_fill_manual(values = c("Mean-only" = col_mean, "Naive (Y=X)" = col_naive, "OLS" = col_ols)) +
  labs(title = "Typical Prediction Error, by Model", x = NULL, y = "Error (cm)") +
  theme_deck
```

OLS makes the smallest errors — it's the best of the three.

## Explanatory Models vs Predictor Models

In an explanatory model , the outcome (y) is known to the investigator. The focus is on solving for the slope. 
In a predictive model, the slope and intercept are known to the investigator (based on training data). The focus is on solving for y. 

## Galton's Own Discovery

The OLS slope, `r round(slope_ols,2)`, is well below 1.

Tall fathers have sons who are tall, but closer to average. Short fathers have sons who are short, but also closer to average.

Galton called this **"regression toward mediocrity"** — where the word *regression* comes from.

## Recap

1.  No predictor → guess the **mean**.
2.  Two rigid rules (mean-only, Y = X) both fall short.
3.  OLS **solves** for the slope and intercept that minimize error.
4.  OLS makes the smallest prediction errors of the three.
5.  Its slope < 1 reveals **regression to the mean** — Galton's original insight.
